---
title: "poputils"
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    fig_caption: yes
    toc: true
    toc_depth: 2
    number_sections: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{poputils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Aims of package

**poputils** contains tools for carrying out tasks that are commonly encountered when working with demographic data. The design strategy of **poputils** is:

- Fit into [tidyverse](https://www.tidyverse.org) workflows. For instance, use data frames for inputs and outputs, use [tidyselect](https://tidyselect.r-lib.org/reference/language.html) methods to specify variables, and follow tidyverse conventions for variable names.
- Accommodate uncertainty, using [rvecs](https://bayesiandemography.github.io/rvec/reference/rvec.html). An rvec is a random vector holding multiple draws that behaves similarly to an ordinary R vector.
- Allow users to work directly with age and time labels, based on common set of methods.

Some functions in **poputils** are aimed directly at data analysts working on demographic datasets. Others are aimed at programmers creating functions to be used at data analysts.


# Tools for data analysts

## Labels

### Age

Producers of demographic data follow a wide variety of styles for labeling age groups. **poputils** contains tools for parsing and manipulating age group labels. 

Age label functions in **poputils** require that age labels belong to one of three types:

- `"single"`. Single years of age, possibly including an open age group, eg `"0", `"81"`, `"17"`, `"100+"`. 
- `"five"`. Five-year age groups, possibly including an open age group, eg `"0-4"`, `"80-84"`, `"15-19"`, `"100+"`.
- `"lt"`. Life table age groups. Like `"five"`, but with the `"0-4"` age group split into `"0"` and `"1-4"`.

Age labels created by **poputils** functions such as `age_labels()` follow a standard set of rules. Many age labels created using other rules can, however, be parsed by **poputils** functions,

```{r}
library(poputils)
library(dplyr, warn.conflicts = FALSE)
tibble(original = c("5 to 9", "5_9", "05-09"),
       reformated = reformat_age(original))
```

Functions `age_lower()`, `age_upper()`, and  `age_mid()` extract information about lower limits, upper limits, and centers of age groups. This can be useful for ordering data
```{r}
df <- data.frame(age = c("5-9", "0-4", "15-19", "10-14"),
                 population = c(3, 7, 2, 4))
df
df |>
  arrange(age_lower(age))
```

and plotting

```{r}
library(ggplot2)
ggplot(df, aes(x = age_mid(age), y = population)) +
  geom_point()
```

among other things.

Functions `combine_age()` and `set_age_open()` can be used to collapse age groups,
```{r}
tibble(age = age_labels("lt", max = 30),
       age_5 = combine_age(age, to = "five"),
       age_25plus = set_age_open(age, lower = 20))
```

The aim is that users should be able to with age group labels throughout the analysis.

### Sex/gender

Function `reformat_sex()` converts sex/gender categories to `"Female"`, `"Male"`, and any additional categories specified through the `except` argument,
```{r}
reformat_sex(c("M", "F", "Diverse", "Fem"), except = "Diverse")
```


## Life tables and life expectancy

A life table a way of summarizing mortality conditions. It consists of quantities calculated from age-specific mortality rates. The most widely-used life table quantity is life expectancy at birth.

### Basic functionality

Life tables can be calculated from age-specific mortality rates using function `lifetab()`. 

```{r}
nzmort |>
  filter(year == 2022,
         gender == "Female") |>
  lifetab(mx = mx)  
```

`lifetab()` and `lifeexp()` both have a `by` argument. Separate results are calculated for each combination of the `by` variables,
```{r}
nzmort |>
  lifeexp(mx = mx,
          by = c(gender, year))  
```

The same effect can be obtained using `dplyr::group_by()`,
```{r}
nzmort |>
  group_by(gender, year) |>
  lifeexp(mx = mx)
```



### Calculation methods

Demographers have developed many methods for converting mortality rates into life table quantities [@preston2001demography; @keyfitz2005applied]. The main challenge is capturing the way that mortality rates vary within each age interval. Mortality rates are expecially variable during the first year of life.

It turns out that, for the purposes of constructing life tables, all relevant information about variation within an age interval can be captured by a single number: the average length of time lived in the interval by people who die in the interval [@preston2001demography, p.43]. This number is denoted $_na_x$, where $x$ is exact age at the start of the internal, and $n$ is the length of the interval. The quantity $_5a_{20}$, for instance, refers to the average number of years lived after their 20th birthday by people who die between their 20th and 25th birthdays. When $n=1$, the $n$ subscript is typically omitted.

Functions `lifetab()` and `lifeexp()` have four arguments for specifying calculation methods: 

- `infant`, which specifies how $a_0$ is calculated,
- `child`, which specifies how $_4a_1$ is calculated,
- `closed`, which specifies how $_na_x$ for all other closed intervals are calculated, and 
- `open`, which specifies how the final interval, $_{\infty}a_x$ is calculated.

Different choices of method are available for each argument. In some cases, different formulas are used for females and males. The choices and formulas are as follows:

| `argument` | `method`         | `sex`      |                                                                                                                                                           |
| :------- | :------------- | :------- | :------------------------                                                                                                                                        |
| `infant` | `"constant"`   | \<any\>    | $$a_0 = \frac{1 - (m_0 + 1) e^{-m_0}}{m_0 (1 - e^{-m_0})}$$                                                                        |
| `infant` | `"linear"`     | \<any\>    | $$a_0 = 0.5$$                                                                                                                                                      |
| `infant` | `"CD"`         | Female   | $$a_0 = \begin{cases} 0.053 + 2.8 m_0 & 0 \le m_0 < 0.107 \\ 0.35 &  m_0 \ge 0.107 \end{cases}$$                                                                   |
| `infant` | `"CD"`         | Male     | $$a_0 = \begin{cases} 0.045 + 2.684 m_0 & 0 \le m_0 < 0.107 \\ 0.33 &  m_0 \ge 0.107 \end{cases}$$                                                                 |
| `infant` | `"AK"`        | Female   | $$a_0 = \begin{cases} 0.14903 - 2.05527 m_0 & 0 \le m_0 < 0.01724 \\ 0.04667 + 3.88089 m_0 & 0.01724 \le m_0 < 0.06891 \\ 0.31411 &  m_0 \ge 0.06891 \end{cases}$$ |
| `infant` | `"AK"`        | Male     | $$a_0 = \begin{cases} 0.14929 - 1.99545 m_0 & 0 \le m_0 <  0.023 \\ 0.02832 + 3.26021 m_0 &  0.023 \le m_0 < 0.08307 \\ 0.29915 &  m_0 \ge 0.08307 \end{cases}$$   |
| `child`  | `"constant"`   | \<any\>    | $$_4a_1 = \frac{1 - (4 \times {_4}m_1 + 1) e^{-4 \times {_4}m_1}}{_4m_1 (1 - e^{-4 \times {_4}m_1})}$$                                                                      |
| `child`  | `"linear"`     | \<any\>    | $$_4a_1 = 2$$                                                                                                                                                      |
| `child`  | `"CD"`         | Female   | $$_4a_1 = \begin{cases} 1.522 - 1.518 m_0 & 0 \le m_0 < 0.107 \\ 1.361 &  m_0 \ge 0.107 \end{cases}$$                                                              |
| `child`  | `"CD"`         | Male     | $$_4a_1 = \begin{cases} 1.651 - 2.816 m_0 & 0 \le m_0 < 0.107 \\ 1.352 &  m_0 \ge 0.107 \end{cases}$$                                                              |
| `closed`   | `"constant"`   | \<any\>    | $$_na_x = \frac{1 - (n \times {_n}m_x + 1) e^{-n \times {_n}m_x}}{_nm_x (1 - e^{-n \times {_n}m_x})}$$ |
| `closed`   | `"linear"`     | \<any\>    | $$_na_x = 0.5 n$$                                                                                                                                                  |
| `open`     | `"constant"`   | \<any\>    | $$_{\infty}a_{\omega} = \frac{1}{_{\infty}m_{\omega}}$$                                                                                                                                |

In the table above, the values for `"CD"` are from @coale1983regional, cited in @preston2001demography; the values for `"AK"` are from @andreev2015average, cited in @wilmoth2021methods; and the values for `"constant"` are expected values for an exponential distribution with rate $_nm_x$ that has been right-truncated at $n$.

Once the $_na_x$ have been determined, the life table is fully specified, and the required calculations can be carried out with no further input from the user.

The probability of dying within each interval is

$$_nq_x = \frac{n \times {_n}m_x}{1 + (n -  {_n}a_x) \times {_nm_x}},$$
with $_{\infty}q_{\omega} = 1$. Quantity $l_x$ is the number of people surviving to exact age $x$. In `lifetab()`, by default, $l_0 = 100,000$. Remaining values are calculated using 

$$l_{x+n} = (1 - {_nq_x}) \times l_x.$$
Quantity $_nd_x$ is the number of people who die between exact ages $x$ and $x+n$,

$$_nd_x = l_x - l_{x+n}.$$

Quantity $_nL_x$ is the number of person-years lived between exact ages $x$ and $x+n$. It consists of person-years lived by people who survive the interval, plus person-years lived by people who die within the interval,

$$_nL_x = l_{x+n} \times n + {_nd_x} \times {_na_x}.$$
Finally, $e_x$, the number of years of life remaining to a person aged exactly $x$, is 
$$e_x = {_nL_x} + {_nL_{x+n}} + \cdots + {_{\infty}L_{\omega}}$$.

Although the results for `lifetab()` and `lifeexp()` do vary with difference choices for `infant`, `child`, or `closed`, the differences are often small,
```{r}
lin <- nzmort |>
  lifeexp(mx = mx,
          by = c(gender, year),
          infant = "linear",
          prefix = "lin")
ak <- nzmort |>
  lifeexp(mx = mx,
          sex = gender,
          by = year,
          infant = "AK", 
          prefix = "hmd")
inner_join(lin, ak, by = c("year", "gender"))
```

### Uncertainty

The examples of life tables and life expectancy so far have all been based on a deterministic input, `mx` column of data frame `nzmort`,

```{r}
nzmort
```

The data frame `nzmort_rvec` instead uses a rvec to represent mortality rates,

```{r}
library(rvec)
nzmort_rvec
```

The `mx` rvec holds 1000 draws from the posterior distribution from a Bayesian model of mortality. The posterior distribution for infant mortality for females in 2021, for instance, has a posterior median of 0.0032, and a 95% credible interval of (0.0028, 0.0037).

If the input to `lifetab()` or `lifeexp()` is an rvec, then the output will be too. Uncertainty about mortality rates is propagated through to quantities derived from these rates.

```{r}
library(rvec)
nzmort_rvec |>
  filter(year == 2022,
         gender == "Female") |>
  lifetab(mx = mx) |>
  select(age, qx, lx)
```


# Tools for developers

**poputils** provides some functions that developers creating packages to be used by demographers may find useful. 

## Labels

`check_age()` and `age_group_type()` can be useful in functions that involve age group labels. `check_age()` performs some basic validity checks, while `age_group_type()` assesses whether a set of labels belongs to type `"single"`, `"five"`, or `"lt"`.

It is often possible to guess the nature of a demographic variable, or of categories within a demographic variable, based on names and labels. Functions `find_var_age()`, `find_var_sexgender()`, `find_var_time()`, `find_label_female()`, and `find_label_male()` help with these sorts of inferences.

## Data manipulation

Function `groups_colnums()` is helpful when implementing [tidyselect](https://tidyselect.r-lib.org/reference/language.html) methods when the data are held in a grouped data frame.

`matrix_to_list_of_cols()` and `matrix_to_list_of_rows()` convert from matrices to lists of vectors. 

`to_matrix()` converts a data frame to a matrix. The data frame potentially has more than two classification variables, and the rows and/or columns of the matrix can be formed from combinations of these variables.

# Future developments

## Definite

- **Stable populations**. Given mortality and fertility profiles, generate the associated stable population.
- **Time labels** Functions for dealing with time labels analogous to the existing ones dealing with age labels. The functions need to allow for one-month and one-quarter periods, and for 'exact times', ie dates.
- **Multiple decrement life tables** Extend `lifetab()` and `lifeexp()` to allow for multiple decrements.
- **Projection accounting** Functions to turn projected demographic rates, and an initial population, into a projected demographic account. Needs flexibility over dimensions included, and needs deterministic and probabilistic versions.
- **TFR** Function to calculation total fertility rates from age-specific fertility rates
- **Age, period, cohort labels** Functions to allocate events to age groups, periods, or cohorts, based on data on dates of event and dates of birth.


## Possible

- **Aggregation function** Using `dplyr::count()`, `dplyr::summarise()`, or `stats::aggregate()` to aggregate counts or rates in a data frame is awkward. Given that this is such a common operation, it might be worthwhile to do a replacement. 




# References
